
---
title: "Data Review"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(finetune)
library(doParallel) 
library(tidymodels)   
library(themis)       
library(vip)          
library(viridis)   
library(parallel)
library(RANN)
library(DMwR2)
library(readxl)
library(stringr)
library(scales)
library(dplyr)
library(pROC) 
library(MLmetrics)
library(tidyverse)
library(forcats)
library(lubridate)
library(ggthemes)
library(ggridges)
library(ggplot2)
library(ggforce)
library(xgboost)
library(Matrix)
library(tictoc)
library(embed)
library(yardstick)
```

## Load The Data

```{r}
df <- read_excel("~/file.xlsx")
head(df)
```
Raw data includes 41 columns.

## Clean The Data

Imbalanced Data:
```{r}
table(df$animal_type)
df %>%
  count(animal_type, sort = TRUE)
```
Since Most Data Divides Between Dogs And Cats We Will Focuse On Those Types.

Keep Dogs And Cats Only,Drop unnaccary columns:
```{r}
df <- df %>% 
  filter(animal_type %in% c("Dog", "Cat")) %>%
  select(-age_upon_outcome,-time_in_shelter,-animal_id_outcome,-date_of_birth,-sex_upon_outcome,-`age_upon_outcome_(days)`,-`age_upon_outcome_(years)`,-age_upon_outcome_age_group,-outcome_datetime,-outcome_month,-outcome_year,-outcome_monthyear,-outcome_weekday,-outcome_hour,-outcome_number,-dob_year,-dob_month,-dob_monthyear,-animal_id_intake,-found_location,-intake_type,-count,-age_upon_intake,-`age_upon_intake_(days)`,-age_upon_intake_age_group,-intake_datetime,-intake_month,-intake_year,-intake_monthyear,-intake_weekday,-intake_hour,-intake_number,-time_in_shelter_days,-outcome_subtype)

dim(df)
```

Null values:
```{r}
colSums(is.na(df))
```
Few null values-dropping rows:
```{r}
df <- na.omit(df) 
colSums(is.na(df))
```
## Adjust columns, Feature engineering

```{r}
table(df$outcome_type)
```

Merging outcome categories to Adopted/Not Adopted:
```{r}
outcome_mapping <- c(
    'Adoption'= "Adopted",       
    'Rto-Adopt'="Adopted" ,
    'Died'= "Not_Adopted",   
    'Disposal'= "Not_Adopted",     
    'Euthanasia'= "Not_Adopted",    
    'Missing'= "Not_Adopted",          
    'Return to Owner'= "DELETE",
    'Transfer'= "DELETE"   
)
df$outcome_type <- outcome_mapping[df$outcome_type]
df <- df[df$outcome_type!="DELETE",]
table(df$outcome_type)
```
Although the resulting classes are still imbalanced, we will handle the imbalance later during model training using smote. 

Turn breed into Mixed/Purebred:
```{r}
df$"mixed_purebred" <- ifelse(grepl("Mix", df$breed) | grepl("/", df$breed), "Mixed" , "Purebred")
df <- df %>% 
  select(-breed)
table(df$"mixed_purebred")
```

Merging different colors into 3 categories:
```{r}
df <- df %>%
  mutate(color_group = case_when(
    grepl("Black", color, ignore.case = TRUE) ~ "Black", 
    grepl("Brown|Tan|Buff|Red|Gold|Sable|Brindle|Merle|Smoke|Chocolate|Fawn|Liver|Agouti|Chocolate Point|Tabby|Tick|Tortie|Torbie|Chocolate
", color, ignore.case = TRUE) ~ "Brown", 
    grepl("White|Cream|Light|Yellow|Gray|Orange|Silver|Apricot|Pink|Blue|Lynx Point|Seal Point|Point|Silver Tabby|Silver|Calico|Calico Point|Tortie Point|Tricolor|Torbie
", color, ignore.case = TRUE) ~ "Light",
    TRUE ~ "Unknown"
  ))
df <- df %>%
  select(-color)
table(df$color_group)


```

Turn health_status into Healthy/Sick-Aged-Pregnant:
```{r}
df <- df %>%
  mutate(health_status = case_when(
    intake_condition == "Normal" ~ "Healthy",   
    TRUE ~ "Sick_Aged_Pregnant"                          
  )) %>%
  select(-intake_condition)
  
table(df$health_status)

```

Turn sex_upon_intake into sex and neutered columns:
```{r}
df <- df %>%
  mutate(
    sex = case_when(
      str_detect(sex_upon_intake, "Male") ~ "Male",  
      str_detect(sex_upon_intake, "Female") ~ "Female",
      TRUE ~ "Unknown" 
    ),
    neutered = ifelse(str_detect(sex_upon_intake, "Neutered|Spayed"), "Neutered_Spayed","Intact")
  ) %>%
  select(-sex_upon_intake)
 
table(df$sex)
table(df$neutered)
```
Although the resulting classes are still imbalanced, we will handle the imbalance later during model training using class weighting. 

Turn animal_type into numeric column:
```{r}
df <- df %>%
  mutate(
    animal_type = case_when(
      str_detect(animal_type, "Dog") ~ "Dog",  
      str_detect(animal_type, "Cat") ~ "Cat",
      TRUE ~ "Unknown" 
    ))
table(df$animal_type)
```

Change "age_upon_intake_(years)" into age_groups, and age_norm:
```{r}
df["age_norm"]=df$`age_upon_intake_(years)`/max(df$`age_upon_intake_(years)`)
df <- df %>%
  mutate(age_group = case_when(
  `age_upon_intake_(years)`  <= 0.5 ~ "Puppy_Kitten",
  `age_upon_intake_(years)`  <= 2 ~"Juvenile",
  `age_upon_intake_(years)`  <= 7 ~"Adult",
  `age_upon_intake_(years)`  > 7 ~ "Senior",
    TRUE ~ "Unknown"
  )) %>%
  select (-`age_upon_intake_(years)`)
df<-df[df$sex!="Unknown",]
table(df$age_group)
```

convert categorical columns to factors:
```{r}
df$outcome_type<-as.factor(df$outcome_type)
df$mixed_purebred<-as.factor(df$mixed_purebred)
df$color_group<-as.factor(df$color_group)
df$animal_type<-as.factor(df$animal_type)
df$health_status<-as.factor(df$health_status)
df$sex<-as.factor(df$sex)
df$neutered<-as.factor(df$neutered)
df$age_group<-as.factor(df$age_group)
```

View the data after changes:
```{r}
str(df)
colSums(is.na(df))
```
Data is cleaned and adjusted

```{r}
write.csv(
  df,
  file = "clean_data.csv",  
  row.names = FALSE,       
  fileEncoding = "UTF-8"    
)
```


## Review the cleaned data

```{r}
summary(df)
```
Most of the data is inherently imbalanced due to the typical characteristics of animals that end up in shelters. 
We address the class imbalance within the outcome_type target variable by incorporating SMOTE directly into our cross-validation training process. 
For these imbalances within the explanatory features, we leverage the inherent robustness of the XGBoost algorithm. Tree-based ensemble methods like XGBoost are generally less sensitive to skewed or imbalanced distributions in predictor variables compared to some other machine learning algorithms.

Review the data:
```{r}
str(df)
```
One number column and the rest are factors, 9 columns, 36ת682 rows.
The primary entities in the data are animals, specifically focusing on Dogs and Cats.
Dogs: 22,394 entities
Cats: 14,630 entities


The dataset includes 9 features, which can be grouped as follows:

Outcome Information:
outcome_type

Animal Characteristics:
animal_type,
mixed_purebred,
color_group,
sex,
neutered,
age_group,
age_norm,

Health Status:
health_status

Null values check:
```{r}
colSums(is.na(df))
```
No null values after cleaning the data.

## Visualization

```{r}
ggplot(df, aes(x = outcome_type, fill = outcome_type)) +
  geom_bar(width = 0.7, show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(title = "Class Balance: Outcome Types",
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)
```
# Observation :

Roughly four-fifths of records end in “Adopted”, confirming a pronounced class imbalance.

```{r}
ggplot(df, aes(x = outcome_type, y = age_norm, fill = outcome_type)) +
  geom_violin(trim = FALSE, alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, colour = "grey20") +
  scale_fill_viridis_d() +
  labs(title = "Age Distribution by Outcome",
       x = NULL, y = "Age (years)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

# Observation :

Adopted animals are markedly younger on average than those not adopted, hinting that age will be a key predictor.

```{r}
df %>% 
  mutate(health_status = fct_lump_n(health_status, 6)) %>%  # keep top 6 statuses
  group_by(health_status, outcome_type) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  group_by(health_status) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(health_status, prop, fill = outcome_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_viridis_d(name = "Outcome") +
  coord_flip() +
  labs(title = "Adoption Share by Health Status",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12)

```

# Observation :

Healthy animals see adoption shares above 90 %, whereas the “Sick / Aged / Pregnant” group drops below 40 %, underscoring the clinical hurdle to placement.

```{r}
plot_colour_rates <- df %>% 
  mutate(color_group = fct_lump_n(color_group, 8)) %>%          # top-8 colours
  count(color_group, outcome_type, name = "n") %>% 
  group_by(color_group) %>% 
  mutate(prop = n / sum(n)) %>%                                 # share within colour
  ungroup() %>% 
  ggplot(aes(y = fct_rev(color_group), x = prop, fill = outcome_type)) +
  geom_col(position = "fill", width = 0.7) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_viridis_d(option = "C", name = "Outcome") +
  labs(
    title = "Adoption Share by Coat Colour",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.ticks.y       = element_blank()
  )

plot_colour_rates
```

# Observation :

ross all major coat colours, adoption shares cluster tightly around 80 %, indicating that colour no longer materially affects placement likelihood.

```{r}


plot_sex_rates <- df %>% 
  mutate(sex_simple = case_when(
    str_detect(sex, "Male")   ~ "Male",
    str_detect(sex, "Female") ~ "Female",
    TRUE                                   ~ "Unknown"
  )) %>% 
  count(sex_simple, outcome_type, name = "n") %>% 
  group_by(sex_simple) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup() %>% 
  ggplot(aes(x = sex_simple, y = prop, fill = outcome_type)) +
  geom_col(position = "fill", width = 0.7) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_viridis_d(option = "D", name = "Outcome") +
  labs(
    title = "Adoption Share by Sex",
    x     = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.ticks.x       = element_blank()
  )

plot_sex_rates

```
# Observation :

Male and female animals display nearly identical adoption rates (~79 %), suggesting sex plays only a minor role in placement decisions.

## Methods & Results

Parallel-processing setup
```{r}
cores <- parallel::detectCores() - 1
registerDoParallel(cores = cores)
```

Train / test split and CV folds
```{r}
set.seed(2025)
split  <- initial_split(df, strata = outcome_type, prop = 0.80)
train  <- training(split)
test   <- testing(split)
folds  <- vfold_cv(train, v = 10, strata = outcome_type)


rec <- recipe(outcome_type ~ ., data = train) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%  
  step_dummy(all_nominal_predictors())            %>%         
  step_zv(all_predictors())                       %>%          
  step_smote(outcome_type)                        %>%          # SMOTE balancing
  step_normalize(all_numeric_predictors())                      

rec_prep <- prep(rec, training = train)
n_preds  <- ncol(juice(rec_prep, all_predictors()))
```

XGBoost spec
```{r}
xgb_spec <- boost_tree(
  trees          = tune(),
  mtry           = tune(),
  tree_depth     = tune(),
  learn_rate     = tune(),
  min_n          = tune(),
  loss_reduction = tune(),
  sample_size    = tune(),
  stop_iter      = 20
) %>%
  set_engine(
    "xgboost",
    objective   = "binary:logistic",
    eval_metric = "auc",
    nthread     = cores
  ) %>%
  set_mode("classification")
```

Hyper-parameter space
```{r}
param_set <- parameters(
  trees(range = c(200L, 1500L)),
  mtry(range  = c(1L, n_preds)),
  tree_depth(),
  learn_rate(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop()
)
```

Workflow
```{r}
wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(xgb_spec)

metric_fun <- metric_set(roc_auc, pr_auc, accuracy, sens, yardstick::spec, f_meas)
options(tidymodels.progress = TRUE)
```

Racing-ANOVA tuning
```{r}
set.seed(2025)
xgb_race <- tune_race_anova(
  wf,
  resamples = folds,
  param_set = param_set,
  metrics   = metric_fun,
  control   = control_race(
    verbose_elim  = TRUE,
    parallel_over = "everything",
    save_pred     = TRUE
  )
)
```

Finalise & evaluate on hold-out test
```{r}
best_params <- select_best(xgb_race, metric = "roc_auc")
final_wf    <- finalize_workflow(wf, best_params)

final_res <- last_fit(final_wf, split, metrics = metric_fun)
```


Confusion Matrix
```{r}
print(
  collect_predictions(final_res) %>%
    conf_mat(outcome_type, .pred_class)
)
```

Variable importance (top-10)
```{r}
trained <- fit(final_wf, train) 
vip(trained$fit$fit,
    num_features = 10,
    geom         = "col",
    mapping      = aes(fill = Importance)) +
  scale_fill_viridis_c(option = "C", direction = -1) +
  labs(
    title = "Top-10 Feature Importances",
    x = NULL, y = "Importance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "none",
    panel.grid.major.y = element_blank()
  )
```

Best hyper-parameters (selected on CV)
```{r}
print(best_params)
```

Hold-out test metrics
```{r}
test_metrics <- collect_metrics(final_res)
print(test_metrics)
```

Extract accuracy
```{r}
acc <- test_metrics %>% filter(.metric == "accuracy") %>% pull(.estimate)
cat(sprintf("\nAccuracy: %.3f  |  Misclassification error: %.3f\n",
            acc, 1 - acc))
```

ROC-AUC
```{r}
preds <- collect_predictions(final_res)
preds$y_true <-ifelse(preds$outcome_type=="Adopted",1,0)

roc_obj <- roc(preds$y_true, preds$.pred_Adopted)
auc_value <- auc(roc_obj)
print(paste("AUC:", round(auc_value, 3)))

roc_df <- data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities
)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "#2196f3", size = 1.2) +
  geom_abline(linetype = "dashed", color = "grey50") +
  annotate("text", x = 0.7, y = 0.2, 
           label = paste("AUC =", round(auc_value, 3)), size = 5, color = "#2196f3") +
  labs(title = "ROC Curve: Adopted vs Not Adopted",
       x = "False Positive Rate (1 - Specificity)",
       y = "True Positive Rate (Sensitivity)") +
  theme_minimal(base_size = 15)
```

XGBoost model summary
```{r}
trained_fit <- pull_workflow_fit(trained) 
print(trained_fit$fit)
```

Additional Graphs
```{r}
df %>%
  count(color_group, health_status, name = "n") %>%      
  group_by(color_group) %>%                               
  mutate(prop = n / sum(n)) %>%                           
  ungroup() %>%
  ggplot(aes(x = color_group, y = prop,
             fill = health_status)) +
  geom_col(width = .7) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Health-status share by coat colour",
       x = "Coat colour", y = "Share of animals") +
  theme_minimal() +
  theme(legend.title = element_blank())

```

## Observation :

Black, brown, and light-coated animals show virtually identical healthy-versus-sick shares, so health status is independent of coat colour.

```{r}
df %>%
  count(color_group, health_status, name = "n") %>%  
  group_by(color_group) %>%                             
  mutate(prop = n / sum(n)) %>%                          
  ungroup() %>%
  ggplot(aes(x = color_group, y = prop,
             fill = health_status)) +
  geom_col(width = .7) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Health-status share by coat colour",
       x = "Coat colour", y = "Share of animals") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

## Observation :

Puppies/kittens, juveniles, adults, and seniors appear in almost the same proportions across every coat colour, confirming age distribution doesn’t vary by coat colour.